Diffusion local times and the time until recombination

### 2015-10-23 notes/diffusion-local-times.Rmd
pde clines

*Useful functions used below:*
```{r setup_fns}
pcline <- function (x,r,recombfn,...) {
    (1/2)*( 1 - sign(x)*recombfn(x,2*r,...) )
}
tanh_cline <- function (x,s,sigma=1) {
    (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_drift <- function (x,s,sigma=1) {
    # log-derivative of tanh_cline
    -(2*sqrt(s)/sigma) * (1-tanh(-2*x*sqrt(s)/sigma)^2) / (1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_het <- function (x,s,sigma=1) {
    # prob of being in a het = (1-s)(1-p)/(1-s(1-p))
    pp <- tanh_cline(x,s,sigma=sigma)
    (1-s)*(1-pp)/(1-s*(1-pp))
}
zeta.reflected <- function (x,r) {
    2*pnorm(abs(x)) - 1 + 2 * exp( r*abs(x) + r^2/2 + pnorm(abs(x)+r,lower.tail=FALSE,log.p=TRUE) )
}
sim_diffusion <- function (drift,tau,nwalks,ntimes,xinit=0,...) {
    # want total time to be tau, so:
    dt <- tau/ntimes
    # put the noise in rws
    rws <- matrix( c( rep_len(xinit,nwalks), # initial conditions
                 rnorm(nwalks*(ntimes-1)) ), nrow=ntimes, ncol=nwalks, byrow=TRUE )
    # and then integrate over it
    for (k in 2:ntimes) {
        rws[k,] <- rws[k-1,]+sqrt(dt)*rws[k,]+dt*drift(rws[k-1,],...)
    }
    attr(rws,"dt") <- dt
    return(rws)
}
weighted_time <- function (rws, weight.fn, use.t=TRUE, ...) {
    colSums( weight.fn(rws[use.t,],...) ) * attr(rws,"dt")
}
clineplot <- function (soln,x=thegrid$x.mid,times=seq(1,nrow(soln),length.out=ntimes),ntimes=20,ylab="probability",ylim=c(0,1),legend=TRUE,...) {
    # plot output of deSolve as clines in a nice way
    graphics::matplot( x, t(soln[times,-1]), type='l', col=rainbow(length(times)), lty=1, 
            xlab="space", ylab=ylab, ylim=ylim, ...)
    if (legend) legend("bottomleft", lty=1, col=rainbow(length(times)), legend=paste("t=",round(soln[times,1],digits=2)), cex=0.5)
    return( invisible( t(soln[times,-1]) ) )
}
```

@@

The key quantity we compute with diffusion theory 
is the probability that a lineage, 
begun at location $x$ and killed at rate $k$ when in heterozygotes,
is still alive at time $\tau$.
Denote this by
$$
\begin{aligned}
    \zeta(x,k,\tau).
\end{aligned}
$$
Now, let $q(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
The probability that a recombination occurs 
between the lineage and the selected locus on either pair of chromatids
is $2r$,
and if this happens, the lineage ends up on either background with probability $1/2$.
Therefore,
for an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probability that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
so if $\zeta = \zeta(x,2r,\tau)$, then
$q=\zeta +(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $q=(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is
$$
\begin{aligned}
    q(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,2r,\tau) \right\} .
\end{aligned}
$$

@@

Actually, the cline gets set up
in a time-inhomogeneous way.
Let's do that, by numerical solution,
then run the lineage back through it.
The forwards-time evolution of the cline,
with isotropic density and migration,
solves the PDE (Barton 1979):
$$\begin{aligned*}
  \partial_t p(x,t) = \frac{\sigma^2}{2} \partial_x^2 p(x,t)+ s p(x,t)(1-p(x,t))(2p(x,t)-1) .
\end{aligned*}$$
```{r fwds_pde, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
yinit <- ifelse( thegrid$x.mid>0, 0, 1 )
tt <- seq(0,tau,length.out=100)
fwds.pde <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, D=1/2, dx=thegrid)$dC
    list( tran + s * y * (1-y) * (2*y-1) )
}
fwds.soln <- ode.1D( y=yinit, times=tt, func=fwds.pde, nspec=1 ) # note FIRST COLUMN IS TIME
image(fwds.soln,legend=TRUE,main="mean cline development", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```
Now, let's run the killed lineage backwards in it:
```{r fwds_back_pde, cache=TRUE, depends="fwds_pde"}
yinit <- rep(1.0,thegrid$N)
cline.interp <- function (T) {
    # return the interpolated cline for a particular time
    which.t <- findInterval(T,fwds.soln[,1],rightmost.closed=TRUE)
    lu <- fwds.soln[c(which.t,which.t+1),1]
    alpha <- (T-lu[1])/diff(lu)
    return( (1-alpha)*fwds.soln[which.t,-1] + alpha*fwds.soln[which.t+1,-1]  )
}
rev.pde <- function (t,y,parms,...) {
    p <- cline.interp(tau-t)
    # p <- setup.prop.1D(grid=thegrid, xy=cbind(thegrid$x.mid,cline.interp(tau-t)), interp="linear")
    tran <- tran.1D(C=y, A=p, D=1/2, dx=thegrid)$dC
    list( tran - r * (1-p) * y )
}
rev.soln <- ode.1D( y=yinit, times=tt, func=rev.pde, nspec=1, tcrit=tau ) # note FIRST COLUMN IS TIME
image(rev.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

@@

Killing at Brownian local times
===============================

@@

Its Laplace transform
=====================

@@

Let's check this agrees with the solution for the PDE.
According to Knight 1981 (Thm 7.4.3), $\zeta'(0) = - 2 r \zeta(0)$;
ReacTran lets you specify the 'flux' at the boundary is proportional to $\zeta(0)$,
where the flux is in our case $\sigma^2/2 \zeta'$.
First note that the solution depends on where we put the boundaries:
```{r reflected_pde_boundaries}
solve.reflected <- function (x.up=-20) {
    x.eps <- 5*eps
    thegrid <- setup.grid.1D(x.up=x.up, x.down=0, N = 5000)
    hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
    yinit <- rep(1.0,thegrid$N)
    tau <- 40
    tt <- seq(0,tau,length.out=20)
    r <- 1
    pde.fn <- function (t,y,parms,...) {
        # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
        tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=4*r)$dC
        list( tran - r * hetfn$mid * y )
    }
    soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
    attr(soln,"x") <- thegrid$x.mid
    return(soln)
}
pde.soln.0 <- solve.reflected(x.up=-5)
pde.soln.1 <- solve.reflected(x.up=-8)
pde.soln.2 <- solve.reflected(x.up=-20)
clineplot(pde.soln.0,x=attr(pde.soln.0,"x"),times=seq_along(tt),xlim=c(-5,0))
clineplot(pde.soln.1,x=attr(pde.soln.1,"x"),add=TRUE)
clineplot(pde.soln.2,x=attr(pde.soln.2,"x"),add=TRUE)
```

And, the PDE solution doesn't agree with the simulations very well: 
clearly, we need a different boundary condition at zero.
This is broken in ReacTran; I've modified the function to fix this.
```{r reflected_pde,cache=TRUE}
source("tran.1D.R")
x.eps <- 5*eps
thegrid <- setup.grid.1D(x.up=-20, x.down=0, N = 5000)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
yinit <- rep(1.0,thegrid$N)
tau <- 40
tt <- seq(0,tau,length.out=20)
r <- 1
pde.fn <- function (t,y,parms,...) {
    # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
    tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=2*r )$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
ps <- clineplot(pde.soln,times=seq_along(tt),xlim=c(-8,0),legend=FALSE)
matlines( thegrid$x.mid, outer( thegrid$x.mid, tt, function(x,t) { zeta.reflected(x/sqrt(t),r*t*x.eps) }), lty=3, col=rainbow(length(tt)) )
legend("bottomleft",lty=1:2,legend=c("PDE","theory"))
```
Still not right...

@@

Here's the cline shape:
```{r show_cline, cache=TRUE}
zeta.pcline <- function (x,r) { pcline(x,r,recombfn=function(x,r){zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau)},tau=1,sigma=1)
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, zeta.pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { zeta.pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```

@@

A better model for the process might be Brownian motion
that, when above zero, has negative drift,
and is killed at some rate.
In other words, 
if $dX_t = -\theta(X_t) dt + dB_t$,
where $\theta(x)=\theta$ for $x \ge 0$, 
and we define
$$\begin{aligned}
\Gamma_t = \text{meas}\{ t : X_t > 0 \} ,
\end{aligned}$$
then we want to know
$$\begin{aligned}
    u(x;\lambda,\theta):=\E^x \left[ \exp\left( - \lambda \Gamma_t \right) \right] .
\end{aligned}$$
(*Small problem:*  $X$ isn't guarenteed to be on the correct side of zero when we want it to be.)
Well, by Girsanov's theorem, we know that
$\E[f(X_t)] = \E[f(B_t)Z(t)]$
where
$$\begin{aligned}
Z(t) = \exp\left[ \int_0^t \theta(B_u) dB_t - \frac{1}{2} \int_0^t \theta(B_u)^2 du \right] 
\end{aligned}$$
and since by Tanaka's formula,
$$\begin{aligned}
\theta \left(\max(0,B_t) - \max(0,B_0)\right) 
&= 
\int_0^t \theta(B_u) du + \theta \ell_t(0) \\
&= 
\theta \Gamma_t + \theta \ell_t(0) ,
\end{aligned}$$
thus
$$\begin{aligned}
Z(t) = \exp\left[ \theta \max(0,B_t) - \theta \max(0,B_0) - \theta \ell_t(0) - \frac{1}{2} \int_0^t \theta(B_u)^2 du \right] 
\end{aligned}$$
and we want to know
$$\begin{aligned}
u(x;\lambda,\theta)
&=
\E^x \left[
\exp\left( 
- \lambda \Gamma_t
\right)
\right]
\end{aligned}$$

>>>>

*Useful functions used below:*
```{r setup_fns}
source("cline-fns.R")
```

@@

The key quantity we compute with diffusion theory 
is the probability that a lineage, 
begun at location $x$ and killed at rate $k$ when in heterozygotes,
is still alive at time $\tau$.
Denote this by
$$
\begin{aligned}
    \zeta(x,k,\tau).
\end{aligned}
$$
Now, let $q(x,r,\tau,A)$ denote the probability that a lineage begun at $x<0$
at a location at recombination distance $r$ from the selected site
inherits from the left-hand side of the barrier
(i.e., an ancestor on $(-\infty,0)$)
$\tau$ generations ago.
The probability that a recombination occurs 
between the lineage and the selected locus on either pair of chromatids
is $2r$,
and if this happens, the lineage ends up on either background with probability $1/2$.
At this point, its probability of originally coming from one or the other background
still depends on its location,
but if we **assume this is 1/2**,
for an individual to the left of the barrier ($x<0$),
the probability of inheriting from the left of the barrier at $\tau$
is the probability that no recombination at the barrier has occurred,
plus one-half the probability that recombination in a heterozygote has occurred,
so if $\zeta = \zeta(x,2r,\tau)$, then
$q=\zeta +(1-\zeta)/2=(1+\zeta)/2$.
If $x>0$ this is the one-half times probability a recombination in a heterozygote has occurred,
i.e., $q=(1-\zeta)/2$.
Combining these,
the probability of ending up on the left-hand side is *approximated* as
$$
\begin{aligned}
    q(x,r,\tau,A) = 
    \frac{1}{2} \left\{ 1 - \sgn(x) \zeta(x,2r,\tau) \right\} .
\end{aligned}
$$

@@

Actually, the cline gets set up
in a time-inhomogeneous way.
Let's do that, by numerical solution,
then run the lineage back through it.
The forwards-time evolution of the cline,
with isotropic density and migration,
solves the PDE (Barton 1979):
$$\begin{aligned*}
  \partial_t p(x,t) = \frac{\sigma^2}{2} \partial_x^2 p(x,t)+ s p(x,t)(1-p(x,t))(2p(x,t)-1) .
\end{aligned*}$$
```{r fwds_pde, cache=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
yinit <- ifelse( thegrid$x.mid>0, 0, 1 )
tt <- seq(0,tau,length.out=100)
fwds.pde <- function (t,y,parms,...) {
    tran <- tran.1D(C=y, D=1/2, dx=thegrid)$dC
    list( tran + s * y * (1-y) * (2*y-1) )
}
fwds.soln <- ode.1D( y=yinit, times=tt, func=fwds.pde, nspec=1 ) # note FIRST COLUMN IS TIME
image(fwds.soln,legend=TRUE,main="mean cline development", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```
Now, let's run the killed lineage backwards in it:
```{r fwds_back_pde, cache=TRUE, depends="fwds_pde"}
yinit <- rep(1.0,thegrid$N)
rev.pde <- function (t,y,parms,...) {
    p <- cline.interp(tau-t,soln=fwds.soln)
    # p <- setup.prop.1D(grid=thegrid, xy=cbind(thegrid$x.mid,cline.interp(tau-t)), interp="linear")
    tran <- tran.1D(C=y, A=p, D=1/2, dx=thegrid)$dC
    list( tran - r * (1-p) * y )
}
rev.soln <- ode.1D( y=yinit, times=tt, func=rev.pde, nspec=1, tcrit=tau ) # note FIRST COLUMN IS TIME
image(rev.soln,legend=TRUE,main="prob of no recomb by t", yaxt='n', ylab='space')
axis(2,at=(pretty(thegrid$x.mid)-min(thegrid$x.mid))/diff(range(thegrid$x.mid)),labels=pretty(thegrid$x.mid))
```

@@

Reflected Brownian motion, killed according to local time at zero
=================================================================

@@

Its Laplace transform
---------------------

@@

Let's check this agrees with the solution for the PDE.
According to Knight 1981 (Thm 7.4.3), $\zeta'(0) = - 2 r \zeta(0)$;
ReacTran lets you specify the 'flux' at the boundary is proportional to $\zeta(0)$,
where the flux is in our case $\sigma^2/2 \zeta'$.
First note that the solution depends on where we put the boundaries:
```{r reflected_pde_boundaries}
source("tran.1D.R")
solve.reflected <- function (x.up=-20) {
    x.eps <- 5*eps
    thegrid <- setup.grid.1D(x.up=x.up, x.down=0, N = 5000)
    hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
    yinit <- rep(1.0,thegrid$N)
    tau <- 40
    tt <- seq(0,tau,length.out=20)
    r <- 1
    pde.fn <- function (t,y,parms,...) {
        # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
        tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=4*r)$dC
        list( tran - r * hetfn$mid * y )
    }
    soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
    attr(soln,"x") <- thegrid$x.mid
    return(soln)
}
pde.soln.0 <- solve.reflected(x.up=-5)
pde.soln.1 <- solve.reflected(x.up=-8)
pde.soln.2 <- solve.reflected(x.up=-20)
clineplot(pde.soln.0,x=attr(pde.soln.0,"x"),times=seq_along(tt),xlim=c(-5,0))
clineplot(pde.soln.1,x=attr(pde.soln.1,"x"),add=TRUE)
clineplot(pde.soln.2,x=attr(pde.soln.2,"x"),add=TRUE)
```

And, the PDE solution doesn't agree with the simulations very well: 
clearly, we need a different boundary condition at zero.
This is broken in ReacTran; I've modified the function to fix this.
```{r reflected_pde,cache=TRUE}
source("tran.1D.R")
x.eps <- 5*eps
thegrid <- setup.grid.1D(x.up=-20, x.down=0, N = 5000)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){abs(x)<x.eps})
yinit <- rep(1.0,thegrid$N)
tau <- 40
tt <- seq(0,tau,length.out=20)
r <- 1
pde.fn <- function (t,y,parms,...) {
    # flux = zeta'(0)/2 = a.bl.down*( zeta(0) - C.down )
    tran <- tran.1D(C=y, A=1, D=1/2, dx=thegrid, C.down=0, a.bl.down=2*r )$dC
    list( tran - r * hetfn$mid * y )
}
pde.soln <- ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
ps <- clineplot(pde.soln,times=seq_along(tt),xlim=c(-8,0),legend=FALSE)
matlines( thegrid$x.mid, outer( thegrid$x.mid, tt, function(x,t) { zeta.reflected(x/sqrt(t),r*t*x.eps) }), lty=3, col=rainbow(length(tt)) )
legend("bottomleft",lty=1:2,legend=c("PDE","theory"))
```
Still not right... hm.

@@

Here's the cline shape:
```{r show_cline, cache=TRUE}
zeta.pcline <- function (x,r) { pcline(x,r,recombfn=function(x,r){zeta.reflected(x=x/sqrt(tau*sigma^2),r=r*tau)},tau=1,sigma=1) }
xx <- seq(-15,15,length.out=500)
for (tau in c(5,50)) {
    plot(xx, zeta.pcline(xx,r=1), type='n', 
            xlab="distance from cline center", ylab="prob inherited from the left", 
            main=as.expression(substitute(tau==mytau,list(mytau=tau))) )
    rr <- seq(0,0.2,length.out=11)
    pp <- sapply( rr, function (r) { zeta.pcline(xx,r=r,tau=tau) } )
    matplot(xx,pp,type='l',lty=1,col=rainbow(length(rr)),add=TRUE)
    legend("topright",lty=1,col=rainbow(length(rr)),legend=paste("r=",rr))
}
```

@@

A better model for the process might be Brownian motion
that, when above zero, has negative drift,
and is killed at some rate.
In other words, 
$$\begin{aligned}
dX_t = -\theta(X_t) dt + dB_t,
\end{aligned}$$
where $\theta(x)=\theta$ for $x \ge 0$, and $\theta(x)=0$ for $x<0$.

Above we simulated the clines deriving from this process.
```{r sim_pushed, cache=TRUE}
rvals <- exp(seq(log(.001),log(0.5),length.out=10))
pushfn <- function(x,s){ifelse(x>0,-4*sqrt(s),0)}
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
tau <- 10
xinit <- seq(-8,8,length.out=nwalks)
sim.rw <- sim_diffusion( pushfn, tau, nwalks, ntimes, s=s, xinit=xinit )
sim.wt <- weighted_time( sim.rw, function (x) { x>0 } )
# distribution of time above zero depending on starting point
plot(xinit,sim.wt,xlab="initial position", ylab="time above zero")
```
We can also get the predicted clines by solving the PDE.
Note that here we are approximating $p(x) = \min(1,\exp(-4x\sqrt{s}))$.
```{r pushed_pde,cached=TRUE}
thegrid <- setup.grid.1D(x.up=-15, x.down=15, N = 5000)
clinefn <- setup.prop.1D(grid=thegrid, func=function(x,s){ifelse(x>0,exp(-4*x*sqrt(s)),1.0)}, s=s)
hetfn <- setup.prop.1D(grid=thegrid, func=function(x){ifelse(x>0,1,0)})
yinit <- rep(1.0,thegrid$N)
tt <- seq(0,tau,length.out=100)
pde.solns <- lapply( rvals, function (r) {
        pde.fn <- function (t,y,parms,...) {
            tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
            if (any(is.na(tran))) { browser() }
            list( tran - r * hetfn$mid * y )
        }
        ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
      } )
layout(t(1:2))
for (k in c(6,10)) {
    clineplot(pde.solns[[k]],main=sprintf("r=%0.2f",tau,rvals[k]))
}
```

Here are the two compared,
by smoothing the probabilities for the simulation:
```{r sim_clines, cache=TRUE, depends=c("sim_pushed","pushed_pde")}
prob.no.recomb <- sapply( rvals, function (r) { fitted(loess( exp(-r*sim.wt) ~ xinit, span=0.1)) } )
matplot(xinit,prob.no.recomb,type='l',ylab="prob of no recomb",
        xlab="space",col=rainbow(length(rvals)),lty=1,main=sprintf("t=%d",tau))
legend("bottomleft",lty=1:5,col=1:6,legend=paste("r=",round(rvals,digits=2)))
for (k in seq_along(rvals)) {
    lines( thegrid$x.mid, pde.solns[[k]][findInterval(tau,pde.solns[[k]][,1]),-1], 
          lty=3, lwd=3, col=rainbow(length(rvals))[k] )
}
```
Well, that looks reasonable, but hard to tell if the discrepancy is due to the loess or not.

And, finally, here's some actual clines predicted by this model:
```{r pushed_cline,cache=TRUE,depends="pushed_pde"}
tau <- 100
tt <- seq(0,tau,length.out=100)
pde.solns <- lapply( rvals, function (r) {
        pde.fn <- function (t,y,parms,...) {
            tran <- tran.1D(C=y, A=clinefn, D=1/2, dx=thegrid)$dC
            if (any(is.na(tran))) { browser() }
            list( tran - r * hetfn$mid * y )
        }
        ode.1D( y=yinit, times=tt, func=pde.fn, nspec=1 ) # note FIRST COLUMN IS TIME
      } )
plot.tvals <- floor(seq(0,tau,length.out=7))[-1]
layout(t(1:2))
for (tval in plot.tvals) {
    matplot( thegrid$x.mid, 
            pcline(thegrid$x.mid, p=sapply( pde.solns, cline.interp, T=tval ) ), 
                   type='l', main=sprintf("t=%d",tval) 
        )
    legend("bottomleft",lty=1:5,col=1:6,legend=paste("r=",round(rvals/2,digits=2)))
}
```

Analytical form
---------------

If we define
$$\begin{aligned}
\Gamma_t = \text{meas}\{ t : X_t > 0 \} ,
\end{aligned}$$
then we want to know
$$\begin{aligned}
    u(x;\lambda,\theta):=\E^x \left[ \exp\left( - \lambda \Gamma_t \right) \right] .
\end{aligned}$$
(*Small problem:*  $X$ isn't guarenteed to be on the correct side of zero when we want it to be.)
Well, by Girsanov's theorem, we know that
$\E[f(X_t)] = \E[f(B_t)Z(t)]$
where
$$\begin{aligned}
Z(t) = \exp\left[ \int_0^t \theta(B_u) dB_t - \frac{1}{2} \int_0^t \theta(B_u)^2 du \right] 
\end{aligned}$$
and since by Tanaka's formula,
$$\begin{aligned}
\theta \left(\max(0,B_t) - \max(0,B_0)\right) 
&= 
\int_0^t \theta(B_u) du + \theta \ell_t(0) \\
&= 
\theta \Gamma_t + \theta \ell_t(0) ,
\end{aligned}$$
thus
$$\begin{aligned}
Z(t) = \exp\left[ \theta \max(0,B_t) - \theta \max(0,B_0) - \theta \ell_t(0) - \frac{1}{2} \int_0^t \theta(B_u)^2 du \right] 
\end{aligned}$$
and we want to know
$$\begin{aligned}
u(x;\lambda,\theta)
&=
\E^x \left[
\exp\left( 
- \lambda \Gamma_t
\right)
\right]
\end{aligned}$$
